### 2016-07-28 part1/lafaySimulatedScenePerception1.tex
update part I with NM comments

\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
\AddressI{$^{1)}$ Institut de Recherche en Communications et  Cybern\'etique de Nantes (IRCCYN), Nantes, France.\\
\hspace*{8pt}gregoire.lafay@irccyn.ec-nantes.fr\\
$^{2)}$ Institut de Recherche et Coordination Acoustique/Musique (IRCAM), Paris, France}

@@

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a  software dedicated to sound arrangement, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points bipolar scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

The study of our sonic environments, both natural and man-made, has gathered growing interest in the past decades, finding applications in fields as varied as context inference and surveillance \cite{heittola13,1540194,park14}, evaluation of urban noise pollution \cite{raimbault_urban_2005} or the emerging area of eco-acoustics \cite{ECOACOUSTICS2014, krause}. The notion of \emph{soundscape}, introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape, is a theoretical framework of choice for that research. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound scene.

Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics or cognitive psychology, but also psycho-linguistics, sociology and data mining.

While the soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}. We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it into distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.

We propose to tackle this issue by introducing a new experimental protocol, in which subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples. The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels. Those annotations enable us to ground the analyses of further perceptive descriptors obtained via standard perceptive evaluations of the simulated scenes, as well as to study the contribution of each sound source.

As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006}.

Only during the 1980s did policy-makers start taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life. To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.

However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $LA_{(eq,V)}$, but needs to take into account other perceptive factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}. ``\,Noise\,''  is a cognitive object which depends upon a listener's appreciation and the context in which it is heard. Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels. Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources. In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds. This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.

The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems. We explore in this paper what additional insights our proposed ``\,generative\,'' approach to soundscape perception analysis can bring to light.

We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.

Results show that 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.

@@

Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception.

The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews). To identify categories, those methods usually require psycho-linguistic and lexical analyses. Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}. Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''. Similar findings were made by \cite{guastavino_ideal_2006}. Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''. Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence.  As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.

Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}. From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}. Several semantic scales show promising results when it commes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}. In addition, \cite{yu2009modeling} shows that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort.  Similar findings are made in \cite{yu2010factors} showing that age and education level significantly influence the sound preference.

Studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment. \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g. belonging to the ``\,human\,'' category. Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level. Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality. Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.

In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within the scene. Acquiring such data may be a serious challenge. From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales. To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli. This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants. Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.

@@

To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds. This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''. The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on conceptual categorization which states that isolated sound classification operates at several levels of abstraction \cite{houix_lexical_2012}. Following this consideration, the dataset is structured along two dimensions:

\begin{itemize}
\item object dimension: where we make the difference between distinct sounds sources such as \textit{voice}, or \textit{car}. 
\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''). 
\end{itemize}

@@

A soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}. Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, mcdermott2013summary}. We thus clearly distinguish between ``\,sound events\,'' and ``\,sound textures\,'' and use two distinct datasets.

@@

\begin{description}
\item[Sample]: a recording of an isolated sound event or texture.
\item[Sound class]: either a collection of samples or a collection of sub-classes, depending of the semantic level of the class.
\item[Track]: a temporal sequence of samples belonging to the same sound class. As samples of a texture track are concatenated seamlessly (see section \ref{sec:simulationProcess}), a texture track can be seen as being made of one sample.
\end{description}

@@

The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space \cite{lafay:hal-01300399}. Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another. This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset. This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.

@@

Following the above described terminology, a soundscape is a sum of tracks. Each track is modeled by a temporal sequence of sound samples belonging to the same sound class. Thus, each track is related to one specific leaf sound class (see \ref{subsection:Sound dataset}). To generate a soundscape, subjects may interact with the track parameters, but not with specific samples; for example, if a subject wishes to put a sequence of car sounds, he may specify how often (on average) cars pass by, and how loud they should be on average, but he cannot control each event individually. This allows to streamline the creation process, and keep the user focused on a global perception of the soundscape.

@@

\begin{enumerate}
\item \textit{Sound level} (dB): for each track, the sound levels of individual samples are drawn from a normal distribution. Subjects may control the sound level of a track by setting the mean and variance of the distribution. 
\item \textit{Global fade In/Out} (sec):  allows the subject to set a fade effect on the whole sequence of samples. 
\item \textit{Start and Stop position} (sec): allows the subject to set the beginning and the end times of a temporal sequence.
\item \textit{Inter onset spacing time} (sec): similarly to the sound level, inter onset spacing times between events are drawn from a normal distribution that may be parametrized by the subject for each event track.
\item \textit{Sample fade In/Out} (sec): allows the subject to set a same fade effect individually on each sample of an event track.
\end{enumerate}

Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}. Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.

@@

We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni). The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}. The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3} run with 10 subjects.

@@

The simulated scenes as well as their annotations are available on an Archive Data Repository \footnote{\url{https://archive.org/details/soundSimulatedUrbanScene}}. Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}.

@@

483 monophonic urban environmental sounds are collected, including 381 sound events and 102 textures are chosen to build the dataset of isolated sound samples. Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder. The remaining come from existing sound banks. All sounds are normalized to the same root mean square (RMS) level.

@@

\subsection{Experiment-1 setup}
\subsubsection*{Task}

@@

\subsection{Set-up experiment-2}
\label{sec:Experiment-2}
\subsubsection*{Task}

>>>>

\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
\AddressI{$^{1)}$ Institut de Recherche en Communications et  Cybern\'etique de Nantes (IRCCYN), Nantes, France.\\
\hspace*{8pt}gregoire.lafay@irccyn.ec-nantes.fr\\
$^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}

@@

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a  software dedicated to sound \gl{edition}, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points bipolar scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

The study of our sonic environments, both natural and man-made, has gathered growing interest in the past decades, finding applications in fields as varied as context inference and surveillance \cite{heittola13,1540194,park14}, evaluation of urban noise pollution \cite{raimbault_urban_2005} or the emerging area of eco-acoustics \cite{ECOACOUSTICS2014, krause}. The notion of \emph{soundscape}, introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape, is a theoretical framework of choice for that research. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation \gl{and perception}, rather than only taking into account the acoustic parameters of the sound scene.

Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptual factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psychophysics or cognitive psychology, but also psycho-linguistics, sociology and data mining.

While the soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{brocolini2012prediction,aletta2016soundscape}. We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it into distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptual measurements.

We propose to tackle this issue by introducing a new experimental protocol, in which subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples. The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels. Those annotations enable us to ground the analyses of further perceptual descriptors obtained via standard perceptual evaluations of the simulated scenes, as well as to study the contribution of each sound source.

As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \gl{NM comments}.

Only during the 1980s did policy-makers start taking into account the link between \gl{``\,community noise\,''} and pollution, considering noise as a significant degradation of the life quality. To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.

However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $L_{Aeq}$, but needs to take into account other perceptual factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}. ``\,Noise\,''  is a cognitive object which depends upon a listener's \gl{assessment} and the context in which it is heard. Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels. Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources. In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds. This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.

The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{kang2006urban,dubois2006cognitive,aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems \cite{ribeiro2013heart}. We explore in this paper what additional insights our proposed ``\,generative\,'' approach to soundscape perception analysis can bring to light.

We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale \gl{NM comments}.

Results show that \gl{NM comments} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.

@@

Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \gl{NM comments}.

The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews). To identify categories, those methods usually require psycho-linguistic and lexical analyses. Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}. Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''. \gl{This distinction is also observed in \cite{guastavino_ideal_2006}}. Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''. Other studies not using \gl{NM comments} the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \gl{NM comments}.  As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.

Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}. From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}. Several semantic scales show promising results when it comes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}. In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}. \gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \gl{NM comments, attention \cite{marquis2005noise} est une review}.

\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment \gl{NM comments, plus développer, ajout article ?}. \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g. belonging to the ``\,human\,'' category \gl{NM comments: eg means exempli gratia (par exemple)}. Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level. Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality. Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.

In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \gl{inside ?} the scene. Acquiring such data may be a serious challenge. From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales. To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli. This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants. Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.

@@

\gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds. This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,'', where the term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}

\gl{The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on the formation of categories of concrete objects. According to Rosch, the category system may be viewed as a two dimensional space, where the ``\,vertical\,'' dimension deals with the inclusiveness of the categories (\emph{e.g.} how are ordered ``\,human\,'', ``\,voice\,'' and ``\,laugh\,'') while the ``\,horizontal\,'' dimension concerns the segmentation of categories lying in the same level of inclusiveness (\emph{e.g.} ``\,voice\,'' and ``\,car\,''). The ``\,vertical\,'' dimension is a taxonomy, where categories are related to one another by means of class inclusion. Rosch formalized three levels of abstraction to structure the ``\,vertical\,'' dimension: superordinate, basic and subordinate. The greater the inclusiveness of a category the higher the abstraction level. Several studies focusing on sound classification  have already used such hierarchical system \cite{houix_lexical_2012}. Following these considerations, our dataset is also structured along two dimensions:}

\begin{itemize}
\item object dimension: where we make the difference between distinct sounds sources such as ``\,voice\,'', or ``\,car\,'' and which is closely related to the ``\,horizontal\,'' dimension of Rosh.
\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''), and which is closely related to the ``\,vertical\,'' dimension of Rosh.
\end{itemize}

@@

\gl{As stated is section \ref{statoftheart}}, a soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}. Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, mcdermott2013summary}. We thus clearly distinguish between ``\,sound events\,'' and ``\,sound textures\,'' and use two distinct datasets.

@@

\begin{description}
\item[Sample]: the recording of an isolated sound event or texture.
\item[Sound class]: either a collection of samples or a collection of sub-classes, depending of the semantic level of the class.
\item[Track]: a temporal sequence of samples belonging to the same sound class. As samples of a texture track are concatenated seamlessly (see section \ref{sec:simulationProcess}), a texture track can be seen as being made of one sample.
\end{description}

@@

The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space, \gl{which was previously tested in an other study} \cite{lafayJAES}. Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another. This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset. This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.

@@

Following the above described terminology, a soundscape is a sum of tracks. Each track is modeled by a temporal sequence of sound samples belonging to the same sound class. Thus, each track is related to one specific leaf sound class (see Section \ref{sec:SoundDataset}). To generate a soundscape, subjects may interact with the track parameters, but not with specific samples; for example, if a subject wishes to put a sequence of car sounds, he may specify how often (on average) cars pass by, and how loud they should be on average, but he cannot control each event individually. This allows to streamline the creation process, and keep the user focused on a global perception of the soundscape.

@@

\begin{enumerate}
\item \textit{Sound level} (dB): for each track, the sound levels of individual samples are drawn from a normal distribution. Subjects may control the sound level of a track by setting the mean and variance of the distribution. 
\item \textit{Global fade In/Out} (sec):  allows the subject to set a fade effect on the whole sequence of samples. 
\item \textit{Start and Stop position} (sec): allows the subject to set the beginning and the end times of a temporal sequence.
\item \textit{Inter onset spacing time} (sec): (event tracks only) similarly to the sound level, inter onset spacing times between events are drawn from a normal distribution that may be parametrized by the subject for each event track.
\item \textit{Sample fade In/Out} (sec): (event tracks only) allows the subject to set a same fade effect individually on each sample of an event track.
\end{enumerate}

Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}. Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\gl{NM comments}

@@

We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni). The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}. The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3}.

%run with 10 subjects.

@@

The simulated scenes as well as their annotations are available on an Archive Data Repository \footnote{\url{https://archive.org/details/soundSimulatedUrbanScene}}. Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}. \gl{NM: Pas d'accord commentaires section IV}

@@

483 monophonic urban environmental sounds are collected, including 381 sound events and 102 textures are chosen to build the dataset of isolated sound samples. Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder. The remaining comes from existing sound banks \gl{NM préciser lesquels ?}. All sounds are normalized to the same root mean square (RMS) level.

@@

\subsection{Set-up Experiment-1}
\subsubsection*{Procedure}

@@

\subsection{Set-up experiment-2}
\label{sec:Experiment-2}
\subsubsection*{Procedure}

Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) to +3 (ideal, very pleasant). The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes. Those ten scenes are repeated at the end of the experiment and the ratings of their first occurrences are not taken into account in the analysis. Subjects listen to at least the 20 first seconds of the scenes and are free to skip the last 10 seconds at their convenience. Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes. \gl{NM comments: la randomization est déjà précisée.}
